Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-08-20 workflow/workflow.Rmd
Update workflow.Rmd
Sandrine: Minor edits to text.

Here, we are applying `RSEC` to the low-dimensional `W` matrix from ZINB-WaVE. To do so, we need to specify the `reducedDim` slot to use with the `reduceMethod` option. Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them. In larger dataset, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).

@@

- `k0s`: the values of `k` to start the sequential algorithm.
- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
- `clusterFunction`: function used for the subsample clustering: `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
- `mergeMethod`: our consensus clustering may find too many small clusters. If that's the case a merging step can be employed to merge clusters that do not show enough differential expression. Here, we skip this step by setting the method as "none".
- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).

@@

The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusters (Figure \@ref(fig:plotcoclust)).

@@

This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object. The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e. a matrix and a vector) or, as below, components of a `SingleCellExperiment` object. When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function. For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.

@@

Here, we show how to use `slingshot` starting from a matrix and a set of cluster labels.

@@

Generalized Additive Models (GAMs) [@gam] are a class of models that replaces the usual linear function of the coefficients with a sum of smooth functions. In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present. This correspond to a simple loess regression, but with the added advantage of providing a framework for the testing of the significance of the association between pseudotime and expression.

@@

Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years. The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach. For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above. This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of each step will be readable by the next step.

@@

Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, datasets made of tens of thousands of cells are now routine. In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells. Hence, there is a growing need for scalable methods.

@@

The authors are grateful to Professor John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Dr. Russell B. Fletcher and Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data. We would also like to thank Michael B. Cole for his contributions to `scone`.

>>>>

Here, we are applying `RSEC` to the low-dimensional `W` matrix from ZINB-WaVE. To do so, we need to specify the `reducedDim` slot to use with the `reduceMethod` option. Here, we skip the merging step to collapse similar clusters based on the amount of differential gene expression between them. In larger datasets, this step may prevent results characterized by a large number of very small clusters (see [@Risso2018] for details).

@@

- `k0s`: the values of `k` to start the sequential algorithm.
- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
- `clusterFunction`: function used for the subsample clustering; `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings, and the number of starting values for the k-means algorithm.
- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
- `mergeMethod`: our consensus clustering may find too many small clusters. If that's the case a merging step can be employed to merge clusters that do not show enough differential expression. Here, we skip this step by setting the method as "none".
- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).

@@

The `plotCoClustering` function produces a heatmap of the co-clustering matrix, which records, for each pair of cells, the proportion of times they were clustered together across the candidate clusterings (Figure \@ref(fig:plotcoclust)).

@@

This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object. The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e., a matrix and a vector) or, as below, components of a `SingleCellExperiment` object. When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` object as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function. For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`, respectively.

@@

Here, we show how to use `slingshot` starting from a matrix of (low-dimensional) expression measures and a set of cluster labels.

@@

Generalized Additive Models (GAMs) [@gam] are a class of models that replaces the usual linear function of the coefficients with a sum of smooth functions. In the code below, we use a special case of the GAM in which only one covariate (pseudotime) is present. This corresponds to a simple loess regression, but with the added advantage of providing a framework for testing the significance of the association between pseudotime and expression.

@@

Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years. The use of the `SingleCellExperiment` class, which is the shared "language" [*** SD: Better term than "language"?] among Bioconductor's scRNA-seq software packages, makes it extremely easy to replace one of the steps above by a different approach. For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global-scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal changes to the code structure above. This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of each step will be readable by the next step.

@@

Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, datasets made of tens of thousands of cells are now routine. Indeed, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells. Hence, there is a growing need for scalable methods.

@@

The authors are grateful to Professor John Ngai (Department of Molecular and Cell Biology, UC Berkeley) and his group members Dr. Russell B. Fletcher and Dr.Diya Das for motivating the research presented in this workflow and for valuable feedback on applications to biological data. We would also like to thank Michael B. Cole for his contributions to `scone`.
